{\bf {\Large Detiding time series of dissolved oxygen for improved estimates of ecosystem metabolism}}

### 2014-08-25 comments.Rnw
complete first draft

To date, only a few analyses have focused on systematic comparisons of estuarine metabolism estimates across broad spatial and temporal scales.  Exceptions include \citet{Caffrey03}, \citet{Caffrey04}, and \citet{Caffrey13}.

>>>>

Results from the simulations and case studies suggested that weighted regression can be a practical approach for detiding \ac{DO} time series to remove the effects of physical advection on estimates of ecosystem metabolism.  However, application of the method may only be appropriate under specific situations.  The case studies were chosen based on the relatively high proportion of metabolism estimates that were anomalous and the strength of correlation between the observed \ac{DO} time series and tidal height.  Despite these similarites among the case studies, detiding had variable effects on metabolism estimates.  The results for Elkhorn Slough and Padilla Bay are of particular concern given that mean annual estimates were substantially different compared to those from the observed \ac{DO} time series.  Although the correlation of \ac{DO} and tidal height was reduced for both cases, in addition to a reduction of anomalous estimates, the relative change in mean metabolism before and after detiding suggests a more careful evaluation of the method is needed.  In particular, alternative window widths should evaluated for the ability to remove tidal effects while preserving the biological signal.  The window widths in the above analysis may have removed variation in the \ac{DO} signal from both of these sources.

Although the above analyses suggest the approach has merit, the case studies emphasize a critical challenge in applying weighted regression to monitoring data. Specifically, the true biological signal is not known and the relative contribution of horizontal advection to bias is not accurately quantified with the available data.  Comparative analyses between systems with varying tidal influence or within-system evaluations of multiple sites at fixed distances are necessary to further validate the performance of weighted regression.  In the absence of additional validation, we propose a precautionary approach for application of the weighted regression to monitoring data.  Weighted regression may be most effective at macrotidal sites with strong evidence of the effects of tidal advection on biological signals.  A weight-of-evidence approach should be used such that the occurrence of anomalous metabolism estimates, strong correlations between observed \ac{DO} and tide height, and clear visual patterns of tide change on \ac{DO} would suggest detiding is appropriate.  The choice of window widths may also produce varying results.  Window widths that produce large changes in mean annual estimates should be interpreted with caution.  In general, a pragmatic approach is emphasized such that results should be evaluated based on the preservation of diel variation from production while exhibiting minimal changes with the tide.  Such an approach, combined with further validation, will support informed management decisions through more accurate estimates of ecosystem metabolism.

### 2014-08-25 discussion.Rnw
complete first draft

The primary objective for development and application of the weighted regression technique was to provide a method for more accurately estimating ecosystem metabolism by removing bias associated with tidal change in observed \ac{DO} time series.  The application of weighted regression to simulated \ac{DO} time series with known characteristics and extension to continuous monitoring data from selected \ac{NERRS} sites provided a proof-of-concept that the method can isolate and remove variation in observed \ac{DO} from tidal change.  Further, aggregation of metabolism estimates using the detided \ac{DO} time series were significantly different than those using the observed data, particularly for relatively long periods of observation. These results suggest that previous estimates of annual means may not accurately reflect true metabolic signals if the effects of tidal variation confound biological signals in observed \ac{DO} time series.  Additionally, variation of mean estimates was substantially reduced for estimates based on detided \ac{DO} time series, suggesting that the certainty of conclusions from detided estimates can be improved even if the mean annual estimates do not change.  Monitoring data for periods of observation less than one year may also produce biased metabolism estimates if observed data are not detided.  Results for each case study showed that significant differences were observed for the detided data at seasonal and monthly aggregations, particularly during summer months for Padilla Bay and Rookery Bay.

Comparisons between detided and biological \ac{DO} time series from the simulations indicated that adequate results can be obtained from the weighted regression for a range of characteristics of \ac{DO} time series, as well as half window widths used in the regression.  An examination of scenarios that produced abnormal results can provide additional insight into factors that affect the performance of weighted regression.  For example, poor performance was observed when the observation uncertainty ($\epsilon_{obs}$) was high and both process uncertainty ($\epsilon_{pro}$) and tidal advection ($DO_{adv}$) were low.  These examples represent time series with excessive random variation, no auto-correlation, and no tidal influence.  Poor performance is expected because the weighted regression models a non-existent tidal signal in a very noisy \ac{DO} time series.  These results were observed even for time series with a large diel component of the biological \ac{DO} signal, suggesting that the model will produce unreliable results in microtidal systems with high noise and no serial correlation.  From a practical perspective, weighted regression should not be applied to noisy time series if there is not sufficient evidence to suggest the variation is related to tidal changes.  Similarly, results with perfect or near-perfect correlations between detided and biological \ac{DO} time series were observed when observation uncertainy and tidal advection effects were not in the simulated time series.  Although there is no logical basis for applying weighted regression to time series with no apparent tidal influences, the results will be as expected, as was true for cases with low tidal advection, high observation uncertainty, and low process uncertainty.  We emphasize that the weighted regression should only be applied to time series for which specific conditions apply, as described below.

The performance metrics used to evaluate weighted regression with the case studies suggested that detiding provided more accuate estimates of ecosystem metabolism.  Correlations of metabolism estimates with tidal height changes after detiding were generally reduced, although trends were not always consistent as correlations were reduced in some cases (Sapelo Island) or reversed in others (Padilla Bay).  However, correlations of net metabolism estimates were reduced in all cases.  Tidal height change provides a proxy measure for horizontal advection that directly affects the measured rate of change of oxygen.  Further, changes in \ac{DO} concentration represent integrated measures of both production and respiration (\cref{metrate}).  The inconsistent results in \cref{tab:cor_res} are potentially related to the effects of horizontal advection on the integrated \ac{DO} signal, given that production and respiration each represent a unique component of the diel \ac{DO} variation that is directly affected by tidal variation.  Regardless, the proportion of anomalous metabolism estimates was reduced by detiding for all case studies, although this measure may also be an incomplete indication of the combined effects of tidal variation.  Negative production and positive respiration estimates suggest assumptions of the open-water method are violated \citep{Needoba12}, whereas `normal' estimates (positive production and negative respiration) may still include a significant source of bias from physical advection by providing over-estimates of true values.  For example, \citet{Nidzieko14} observed that net metabolism at Elkhorn Slough was more often heterotrophic during maximum spring tides that occurred at nighttime, as a substantially larger area of salt marsh was inundated leading to higher respiration estimates.  Although this result supports our general conclusions, a broader discussion regarding whether or not this represents a bias in metabolism from physical advection may be needed.

A strength of the weighted regression approach is the lack of assumptions for describing the relationships between \ac{DO} and tidal variation over time.  Although the functional form of the model is a simple linear regression with two explanatory variables (\cref{funform}), the moving window approach combined with the adaptive weighting scheme allows for quantification of complex tidal effects that may not be possible using alternative approaches.  For example, \citet{Batt12} described the use of a moving window or Kalman filter to improve estimates of ecosystem metabolism in lakes.  The approach minimizes the influences of process and observation uncertainty on observed \ac{DO} time series, which is attributed to the effects of water movement on metabolic signals.  Although a similar approach may be useful for estuaries if process and observation uncertainty are the only sources of variation in the \ac{DO} series, the weighted regression approach is more appropriate if tidal advection is the primary confounding factor.  Additionally, results from the case studies illustrated the ability of the weighted regression approach to model changes over time in the relationships between tidal change and \ac{DO}.  Results for Padilla Bay and Rookery Bay suggested that detiding had the largest effect during the summer, whereas the results for cooler months were not significantly different from the observed.  The weighted regression method produced detided time series that accommodated seasonal variation in \ac{DO} conditional on tidal height change, whereas moving window filters or standard regression techniques would likely not have characterized these dynamic relationships.

>>>>

The primary objective for development and application of the weighted regression technique was to provide a method for more accurately estimating ecosystem metabolism by removing variation associated with tidal change in observed \ac{DO} time series.  The application of weighted regression to simulated \ac{DO} time series with known characteristics and extension to continuous monitoring data from selected \ac{NERRS} sites provided a proof-of-concept that the method can isolate and remove variation in observed \ac{DO} from tidal change.  Further, aggregation of metabolism estimates using the detided \ac{DO} time series were significantly different than those using the observed data, particularly for relatively long periods of observation. These results suggest that previous estimates of annual means may not accurately reflect true metabolic signals if the effects of tidal variation confound biological signals in observed \ac{DO} time series.  Additionally, variation of aggregated metabolism estimates were substantially reduced after detiding, suggesting greater confidence in interpreting detided estimates even if the mean values are similar.  Monitoring data for periods of observation up to one year may also produce significantly different metabolism estimates if observed data are not detided.

Comparisons between detided and biological \ac{DO} time series from the simulations indicated that adequate results can be obtained from weighted regressions for a range of characteristics of \ac{DO} time series.  An examination of scenarios that produced abnormal results can provide additional insight into factors that affect the performance of weighted regression.  For example, poor performance was observed when the observation uncertainty ($\epsilon_{obs}$) was high and both process uncertainty ($\epsilon_{pro}$) and tidal advection ($DO_{adv}$) were low.  These examples represent time series with excessive random variation, no auto-correlation, and no tidal influence.  Poor performance is expected because the weighted regression models a non-existent tidal signal in a very noisy \ac{DO} time series.  These results were observed even for time series with a large diel component of the biological \ac{DO} signal, suggesting that the model will produce random results in microtidal systems with high noise and no serial correlation.  From a practical perspective, weighted regression should not be applied to noisy time series if there is not sufficient evidence to suggest the variation is related to tidal changes.  Similarly, results with perfect or near-perfect correlations between detided and biological \ac{DO} time series were observed when observation uncertainy and tidal effects were not components of the simulated time series.  Although there is no logical basis for applying weighted regression to time series with no apparent tidal influences, the results will be as expected, as was true for cases with low tidal advection, high observation uncertainty, and low process uncertainty.  We emphasize that the weighted regression should only be applied to time series for which specific conditions apply, as described below.

The performance metrics used to evaluate weighted regression with the case studies suggested that detiding provided more accuate estimates of ecosystem metabolism.  Correlations of metabolism estimates with tidal height changes after detiding were generally reduced, although trends were not always consistent as correlations were reduced in some cases (Sapelo Island) or reversed in others (Padilla Bay).  However, correlations of net metabolism estimates were reduced in all cases.  Tidal height change provided a proxy for horizontal advection that directly affects the measured rate of change of oxygen.  Further, changes in \ac{DO} concentration represent integrated measures of both production and respiration (\cref{metrate}).  The inconsistent results in \cref{tab:cor_res} are potentially related to the effects of horizontal advection on the integrated \ac{DO} signal, given that production and respiration each represent a unique component of the diel \ac{DO} variation that is directly affected by tidal variation.  Regardless, the proportion of anomalous metabolism estimates was reduced by detiding for all case studies, although this measure may also be an incomplete indication of the combined effects of tidal variation.  Negative production and positive respiration estimates suggest assumptions of the open-water method are violated \citep{Needoba12}, whereas `normal' estimates (positive production and negative respiration) may still include a significant source of bias from physical advection by providing over-estimates of true values.  For example, \citet{Nidzieko14} observed that net metabolism at Elkhorn Slough was more often heterotrophic during maximum spring tides that occurred at nighttime, as a substantially larger area of salt marsh was inundated leading to higher respiration estimates.  Although this result supports our general conclusions, a broader discussion regarding whether or not this represents an actual bias in metabolism from physical advection may be needed.

A strength of the weighted regression approach is a lack of assumptions describing the relationships between \ac{DO} and tidal variation over time.  Although the functional form of the model is a simple linear regression with only two explanatory variables (\cref{funform}), the moving window approach combined with the adaptive weighting scheme allows for quantification of complex tidal effects that may not be possible using alternative approaches.  For example, \citet{Batt12} described the use of a moving window or Kalman filter \citep{Harvey89} to improve estimates of ecosystem metabolism in lakes.  The approach minimizes the influences of process and observation uncertainty on observed \ac{DO} time series, which is attributed to the effects of water movement on metabolic signals.  Although a similar approach may be useful for estuaries if process and observation uncertainty are the only sources of variation in the \ac{DO} series, the weighted regression approach is more appropriate if periodic tidal advection is the primary confounding factor.  Additionally, results from the case studies illustrated the ability of the weighted regression approach to model changes over time in the relationships between tidal change and \ac{DO}.  Results for Padilla Bay and Rookery Bay suggested that detiding had the largest effect during the summer, whereas the results for cooler months were not significantly different from the observed.  The weighted regression method produced detided time series that accommodated seasonal variation in \ac{DO} conditional on tidal height change, whereas moving window filters or standard regression techniques would likely not have characterized these dynamic relationships.

### 2014-08-25 intro.Rnw
complete first draft

Metabolism estimates from time series of dissolved oxygen provide integrated evaluations of trophic state in aquatic ecosystems \citep{Kemp12,Needoba12}.  Autotrophic or productive systems are characterized by high rates of energy transfer between trophic levels leading to accumulation organic matter, whereas heterotrophic systems are sinks of organic matter that are supported by allochthonous sources of energy input.  Integrated measures of metabolism describe the balance between production and respiration that accounts for varying rates in processes that create and consume organic matter.  Although metabolic rates vary naturally at multiple spatiotemporal scales \citep{Ziegler98,Caffrey04,Russell07}, anthropogenic stressors are generally accepted as contributing factors that increase rates of production \citep{Diaz08}.  Inputs of limiting nutrients beyond background concentrations may decrease the resilience of an ecosystem such that higher rates of production are coupled with higher biological oxygen demand \citep{Yin04,Kemp09}.  Cultural eutrophication is frequently linked to declines in water quality through lower levels of dissolved oxygen, degradation in aquatic vegetation habitat, and increased frequency of harmful algal blooms \citep{Cloern96,Short96,Rabalais02}.  Reliables estimates of ecosystem metabolism are critical for measuring both background rates of production and potential impacts of human activities on ecosytem condition.

Ecosystem metabolism can be estimated using two basic techniques, each of which is appropriate under different conditions or assumptions \citep{Kemp12}.   Bottle-based techniques rely on rate measurements from discrete water quality samples, whereas open-water techniques infer metabolic rates using \textit{in situ} measurements from continuous monitoring data.  Bottle-based techniques are useful for direct partitioning of metabolic contributions into discrete habitats, such as planktonic production rates during specific time periods \citep{Kemp12}.  However, such measurements may be inappropriate for evaluating whole ecosystem metabolism if significant production occurs in habitats that are not sampled, such as benthic or seagrass production.  As such, the open-water technique provides an integrative measure of metabolism by inferring process rates from \textit{in situ}, continuous monitoring data.  Originally for use in streams \citep{Odum56}, the method has been used with varying success in lakes \citep{Staehr10,Coloso11,Batt12} and estuaries \citep{Caffrey04,Hopkinson05,Caffrey13}.  As with any method, the ability of the open-water technique to accurately estimate whole system metabolism depends on whether the assumptions for its use are met.  Such assumptions are are often only implicity verified in practice, leading to potential biases.  The increasing availability of high resolution monitoring data further necessitates more quantitative assessments of the open-water technique.

The open-water method uses the diel fluctuation of dissolved oxygen to infer rates of ecosytem metabolism, after correcting for losses or gains through air-water exchange \citep{Kemp12}.  Daily integrated measurements of metabolism are based on the balance between daytime estimates of gross production and nighttime estimates of respiration extrapolated to a 24 hour period.  The fundamental assumption of the open-water method is that measurements come from a water mass that has the same recent history \citep{Needoba12}.  Estimates of metabolism from a single location may be inaccurate if substantial variation in water column mixing occurs throughout the period of observation \citep{Russell07}.  As such, the original technique designed for use in streams requires the comparison of data from an upstream and downstream station \citep{Odum56}.  Application of the method to systems without continuous flow, such as lakes or estuaries, have often assumed that a single sampling station provides sufficient data for estimating metabolism \cite{Staehr10}.  While single stations may be valid under specific conditions, numerous studies have shown that the open-water method may be inappropriate given the effects of physical mixing \citep{Ziegler98,Caffrey03,Coloso11,Batt12,Nidzieko14}.

The open-water method has been applied to coastal ecosystems with mixed success.  An exhaustive analysis by \citet{Caffrey03} applied the method to estimate metabolim at 28 continuous monitoring stations at 14 US estuaries.  Data from two of the reserves were used to evaluate the assumption of homogeneity of water masses measured by each sensor.  Although significant differences were not observed for metabolism estimates between adjacent stations, the analysis was based on a comparison of means using conventional significance tests rather than a systematic comparison of time series.  Moreover, a portion of metabolism estimates from all stations were negative for production during the day and positive for respiration during the night.  These values were opposite in sign than expected since production increases oxygen during the day (i.e., positive effect on metabolism) and respiration consumes oxygen at night (i.e., negative effect on metabolism).  These `anamolous' values were attributed to violations in the assumption of water-column heterogeneity, particularly in relation to tidal advection.

The effects of tidal advection on estimates of ecosystem metabolism have been a point of concern in numerous studies \citep{Monbet92,Ziegler98,Caffrey03,Collins13,Howarth14}, although systematic estimates of its effects in tidal systems have been minimal.  An exception is \citet{Nidzieko14} that described quantitative assessment of the effects of fortnightly tidal modulations on metabolism estimates.  Using a control volume approach to measure fluxes into and out of a shallow tidal creek, significant biases in metabolism estimates were observed.  Net heterotrophy was observed during spring tides, whereas metabolism was balanced during neap tides.  The timing of irradiance relative to the tidal cycle was a primary factor contributing to heterotrophy during summer months such that maximum tides occurred during the night, increasing total area for respiration.  The results of the analysis, although specific to the study location, suggest that the effects of tidal advection on \ac{DO} measurements are of primary concern when selecting locations and length of time for sonde deployment in estuaries.  In many cases, the relative magnitude of these effects may be a significant source of bias.  Analytical techniques to evaluate and correct for tidal advection could improve certainty in metabolism estimates and increase the use of data from shorter deployment periods if sources of bias are quantified and removed.

This article describes the theory and application of a novel method for detiding an observed \ac{DO} time series to more accurately quantify estimates of ecosystem metabolim for estuaries.  Specifically, the method characterizes the effects of tidal advection on \ac{DO} observations to improve estimates of open-water metabolism with multi-year time series of high frequency ($<$ one hour) water quality data.  The focus of our analysis is the use of a weighted regression method previously developed for trend analysis of pollutant concentrations in streams and rivers \citep{Hirsch10} and recently adapted for trend evaluation of water quality in estuaries \citepalias{BeckIR}.  The weighted regression approach creates dynamic predictions of \ac{DO} as a function of time and tidal height change, which are then used to normalize, or detide, the \ac{DO} signal.  First, we use a simulation approach to create time series of \ac{DO} observations with known characteristics to evaluate ability of the weighted regression to remove the effects of tidal advection.  Second, four case studies from the \aclu{NERRS} (\acs{NERRS}, \citealt{Wenner04}) are used to validate the method for improving estimates of ecosystem metabolism.  Overall, the analysis is meant to improve the certainty of information obtained from monitoring data by evaluating the relative roles of biological and physical processes in estuarine systems.  Applications are expected to have implications for ecosystem management by outlining strategies for interpreting water quality data with more accuracy.

>>>>

Metabolism estimates from time series of dissolved oxygen provide integrated estimates of production in aquatic ecosystems \citep{Kemp12,Needoba12}.  Autotrophic or productive systems are characterized by high rates of energy transfer between trophic levels leading to accumulation of organic matter, whereas heterotrophic systems are sinks of organic matter that are supported by allochthonous sources of energy input.  Integrated measures of metabolism describe the balance between production and respiration that accounts for varying rates in processes that create and consume organic matter.  Although metabolic rates vary naturally at multiple spatiotemporal scales \citep{Ziegler98,Caffrey04,Russell07}, anthropogenic stressors are often contributing factors that increase rates of production \citep{Diaz08}.  Inputs of limiting nutrients beyond background concentrations may decrease the resilience of an ecosystem such that higher rates of production are coupled with higher biological oxygen demand \citep{Yin04,Kemp09}.  Cultural eutrophication is frequently linked to declines in water quality through lower levels of dissolved oxygen, degradation in aquatic vegetation habitat, and increased frequency of harmful algal blooms \citep{Cloern96,Short96,Rabalais02}.  Reliables estimates of ecosystem metabolism are critical for measuring both background rates of production and potential impacts of human activities on ecosytem condition.

Ecosystem metabolism can be estimated using two basic techniques, each of which is appropriate under different conditions or assumptions \citep{Kemp12}.   Bottle-based techniques rely on rate measurements from discrete water quality samples, whereas open-water techniques infer metabolic rates using \textit{in situ} measurements from continuous monitoring data.  Bottle-based techniques are useful for partitioning metabolic contributions into discrete habitats, such as pelagic production rates during specific time periods \citep{Kemp12}.  However, such measurements may be insufficient for evaluating whole ecosystem metabolism if significant production occurs in habitats that are not sampled, such as the benthos.  As such, the open-water technique provides an integrative measure of metabolism by inferring process rates from \textit{in situ}, continuous monitoring data.  Originally for use in streams \citep{Odum56}, the method has been used with varying success in lakes \citep{Staehr10,Coloso11,Batt12} and estuaries \citep{Caffrey04,Hopkinson05,Caffrey13}.  As with any method, the ability of the open-water technique to accurately estimate whole system metabolism depends on whether the assumptions for its use are not violated.  Such assumptions are often only implicity verified in practice, leading to potential biases.

The open-water method uses the diel fluctuation of dissolved oxygen to infer rates of ecosytem metabolism, after correcting for losses or gains through air-water exchange \citep{Kemp12}.  Daily integrated measurements of metabolism represent the balance between daytime estimates of gross production and nighttime estimates of respiration.  The fundamental assumption is that \ac{DO} measurements describe the same water mass over time \citep{Needoba12}.  Estimates of metabolism from a single location may be inaccurate if substantial variation in water column mixing occurs throughout the period of observation \citep{Russell07}.  Application of the technique to estimate metabolism in streams requires the comparison of data from an upstream and downstream station \citep{Odum56}.  Application to lakes or estuaries have often assumed that a single sampling station provides sufficient data for estimating metabolism \cite{Staehr10}.  While single stations may be valid under specific conditions, numerous studies have shown that the open-water method may be inappropriate given the effects of physical mixing in lakes or estuaries \citep{Ziegler98,Caffrey03,Coloso11,Batt12,Nidzieko14}.  An exhaustive analysis by \citet{Caffrey03} applied the open-water method to estimate metabolim at 28 continuous monitoring stations at 14 US estuaries.  A non-trivial portion of the estimates were negative for production during the day and positive for respiration during the night.  These `anamolous' values were attributed to the effects of tidal advection such that water masses with different metabolic histories were sampled at multiple sites.

The effects of tidal advection on estimates of ecosystem metabolism have been a point of concern in numerous studies \citep{Monbet92,Ziegler98,Caffrey03,Collins13,Howarth14}, although systematic estimates of its effects have been minimal.  Analytical techniques to evaluate and correct for tidal advection could improve certainty in metabolism estimates and increase the use of data from shorter deployment periods if sources of bias are removed.  This article describes the theory and application of a method for detiding an observed \ac{DO} time series to more accurately quantify estimates of ecosystem metabolim for estuaries.  Specifically, the method characterizes the effects of tidal advection on \ac{DO} observations to improve estimates of open-water metabolism with multi-year time series of high frequency ($<$ one hour) water quality data.  The focus of our analysis is the use of a weighted regression method previously developed for trend analysis of pollutant concentrations in streams and rivers \citep{Hirsch10} and recently adapted for trend evaluation of water quality in estuaries \citepalias{BeckIR}.  The weighted regression approach creates dynamic predictions of \ac{DO} as a function of time and tidal height change, which are then used to normalize, or detide, the \ac{DO} signal.  First, we use a simulation approach to create time series of \ac{DO} observations with known characteristics to evaluate ability of the weighted regression to remove the effects of tidal advection.  Second, four case studies from the \aclu{NERRS} (\acs{NERRS}, \citealt{Wenner04}) are used to validate the method for improving estimates of ecosystem metabolism.  Overall, the analysis is meant to improve the certainty of information obtained from monitoring data by evaluating the relative roles of biological and physical processes in estuarine systems.  Applications are expected to have implications for ecosystem management by outlining strategies for interpreting water quality data with more accuracy.

### 2014-08-25 mat_proc.Rnw
complete first draft

The weighted regression model for detiding \ac{DO} time series was adapted from the \ac{WRTDS} method developed by \citet{Hirsch10}.  The \ac{WRTDS} method was developed to model pollutant concentration in streams and normalize predictions to changes in discharge.  The functional form of our model is as follows:
\begin{equation}\label{funform}
DO_{obs}= \beta_0 + \beta_1 t + \beta_2 H
\end{equation}
where observed \ac{DO} is a linear function of decimal time $t$ and astronomical tidal height $H$. Decimal time is a continuous variable for the day and time of each obervation with time as a proportion of the number of total observations added to each day.  The beginning of each day was considered the nearest thirty minute observation (i.e., on the hour and half hour) at which sunrise was expected for a given location and time of year.  Days were centered on the diel cycle rather than conventional times given that the objective was to develop a predictive model relevant for biological \ac{DO} variation that follows solar and seasonal cycles.  The functional form also differed from the original \ac{WRTDS} method that included parameters to estimate variation of the response variable on a sinuisoidal period.  \ac{DO} variation was not modeled using this approach because rates of change may be abrupt following diurnal variation in irradiance or daily \ac{DO} variation may be muted given the weather, as on cloudy days.

Weighted regression is implemented as a moving window that allows for estimation of \ac{DO} throughout the time series by adapting to variation through time as a function of tide. Regression models are estimated sequentially for each observation in the time series using dynamic weight vectors that change with the center of the window.  Weight vectors quantify the relevance of observations to the center of the window in respect to decimal time, hour of the day, and tidal height.  Specifically, weights are assigned using a tri-cube weighting function \citep{Hirsch10}:
\begin{equation}
w= \left\{ 
  \begin{array}{l l}
    \left(1-\left(d/h\right)^3\right)^3 & \quad \textrm{if } |d| \leq h \\
    0 & \quad \textrm{if } |d| > h 
  \end{array} \right.
\end{equation}
where the weight $w$ of each observation is inversely proportional to the distance $d$ from the center of the window such that observations more similar to the point of reference are given higher importance in the regression.  Weights exceeding the maximum width of the window $h$ are equal to zero.  The tri-cube weighting function is similar to a normal distribution such that weights decrease gradually from the center until the maximum window width is reached.  Observations that are half the distance from the center of the window to the maximum window width are weighted one third less than values at the center.  Regressions that use simpler windows (e.g., boxcar approach) are  more sensitive to influential observations as they enter or leave the window, whereas the tri-cube function minimizes their effect through gradual weighting of observations from the center \citep{Hirsch10}.  The weight vector for each observation is the product of three separate weight vectors for decimal time, hour, and tidal height. A low weight is given to an observation if any of the three weighting values were not similar to the center of the window since the final weight vector is the product of three weight vectors for each variable (see the link in the \hyperref[multi]{multimedia} section for graphical display of different weights).

A nontrivial issue with weighted regression is the choice of window width for calculating weights.  Excessively large or small window widths may respectively under- or over-fit the data.  Additionally, optimal window widths may depend on the objective for using the model.  The weighted regression approach can be used for both predicting \ac{DO} and normalizing to remove the variance in the \ac{DO} signal from tidal changes.  Optimal window widths that minimize prediction error or fit to the observed data are typically smaller than the optimum window widths for normalizing the time series.  Similarly, window widths that more effectively detide the \ac{DO} signal may produce predictions for the observed data that are not optimal.  Evaluations of the weighted regression method with simulated \ac{DO} time series, described below, used different window widths to identify an approximate optimal window width for detiding the \ac{DO} signal.  As such, the ability of the models to predict observed \ac{DO} was not a primary concern given that the optimal window width for detiding likely corresponds to a model that predicts \ac{DO} as a function of tide rather than observed \ac{DO} as a function of both tide and biological variation.

@@

The primary objective of the analysis was to evaluate ability of the weighted regression method to detide a \ac{DO} signal to obtain more accurate estimates of metabolism.  \citet{Hirsch10} developed the normalization approach for the \ac{WRTDS} method using a two-dimensional interpolation grid that contains predicted values of pollutant concentrations across the time series and the range of stream discharge values observed in the study system \citep{Hirsch10}.  Normalized values for pollutant concentration are obtained by averaging the model predictions across the discharge values that are likely to occur on a given day to provide an estimate that is independent of flow variation.

Predicted values of \ac{DO} concentration were normalized to remove variation from tidal height changes, although the approach herein differs slightly from \citet{Hirsch10}.  Our approach uses weighted regression to isolate sources of variation in the observed \ac{DO} signal that are related to unique effects of tidal height and biological process (\cref{fig:do_dtd}).  Two sets of values are predicted for the observed time series $DO_{obs}$, rather than creating an interpolation grid.  The first set of values uses the tidal height of an observation and second set uses the mean tidal height across the time series, $DO_{tid}$ and $DO_{nrm}$ respectively.  In other words, the first set of predictions represent \ac{DO} as a function of time and tide, where the second set represents \ac{DO} conditional on time and mean tidal height:
\begin{equation} \label{do_tid}
DO_{tid} = f(DO_{obs}|H, t)
\end{equation}
\begin{equation} \label{do_nrm}
DO_{nrm} = f(DO_{obs}|\bar{H}, t)
\end{equation}

>>>>

The weighted regression model for detiding \ac{DO} time series was adapted from the \ac{WRTDS} method developed by \citet{Hirsch10}.  The functional form of our model is a simple linear regression:
\begin{equation}\label{funform}
DO_{obs}= \beta_0 + \beta_1 t + \beta_2 H
\end{equation}
where observed \ac{DO} is a linear function of decimal time $t$ and astronomical tidal height $H$. Decimal time is a continuous variable for the day and time of each obervation with time as a proportion of the number of total observations added to each day.  The beginning of each day was considered the nearest thirty minute observation at which sunrise was expected for a given location and time of year.  Days were centered on the diel cycle rather than conventional times (i.e., midnight) given that the objective was to develop a predictive model relevant for biological \ac{DO} variation that follows solar and seasonal cycles.  The functional form also differed from the original \ac{WRTDS} method that included parameters to estimate variation of the response variable on a sinuisoidal period.  \ac{DO} variation was not modeled using this approach because rates of change may be abrupt following diurnal variation in irradiance or daily \ac{DO} variation may be muted given the weather, as on cloudy days.

Weighted regression is implemented as a moving window that allows for estimation of \ac{DO} throughout the time series by adapting to variation through time as a function of tide. Regression models are estimated sequentially for each observation in the time series using dynamic weight vectors that change with the center of the window.  Weight vectors quantify the relevance of observations to the center of the window in respect to decimal time, hour of the day, and tidal height.  Specifically, weights are assigned to each variable using a tri-cube weighting function \citep{Tukey77,Hirsch10}:
\begin{equation}
w= \left\{ 
  \begin{array}{l l}
    \left(1-\left(d/h\right)^3\right)^3 & \quad \textrm{if } |d| \leq h \\
    0 & \quad \textrm{if } |d| > h 
  \end{array} \right.
\end{equation}
where the weight $w$ of each observation is inversely proportional to the distance $d$ from the center of the window such that observations more similar to the point of reference are given higher importance in the regression.  Weights exceeding the maximum width of the window $h$ are equal to zero.  The tri-cube weighting function is similar to a normal distribution such that weights decrease gradually from the center until the maximum window width is reached.  Observations that are half the distance from the center of the window to the maximum window width are weighted one third less than values at the center.  Regressions that use simpler windows (e.g., boxcar approach) are  more sensitive to influential observations as they enter or leave the window, whereas the tri-cube function minimizes their effect through gradual weighting of observations from the center \citep{Hirsch10}.  The final weight vector for each observation is the product of three separate weight vectors for decimal time (day), hour, and tidal height. Windows for decimal time and hour describe the the amount of time from the center of window, whereas the window for tidal height describes the distance as a proportion of the total range.  A low weight is given to an observation if any of the three weighting values were not similar to the center of the window since the final weight vector is the product of three weight vectors for each variable (see the link in the \hyperref[multi]{multimedia} section for graphical display of different weights).

A nontrivial issue with weighted regression is the choice of window width for each weight vector.  Excessively large or small window widths may respectively under- or over-fit the data.  Additionally, optimal window widths may depend on the objective for using the model.  The weighted regression approach can be used for predicting \ac{DO} or normalizing to remove the variance from tidal changes.  Optimal window widths that minimize prediction error or fit to the observed data are typically smaller than the optimum window widths for normalizing the time series.  Similarly, window widths that more effectively detide the \ac{DO} signal may produce predictions for the observed data that are not optimal.  Evaluations of the weighted regression method with simulated \ac{DO} time series, described below, used multiple window widths to identify an approximate optimal window width for detiding the \ac{DO} signal.  As such, the ability of the models to predict observed \ac{DO} was not a primary concern given that the optimal window width for detiding likely corresponds to a model that predicts \ac{DO} as a function of tide rather than observed \ac{DO} as a function of both tide and biological variation.

@@

The primary objective of the weighted regression is to detide a \ac{DO} signal to obtain more accurate estimates of metabolism.  \citet{Hirsch10} developed the normalization approach for the \ac{WRTDS} method to remove the effects of stream discharge on estimates of pollutant concentration using a two-dimensional interpolation grid of predicted pollutant concentrations across the time series and the range of stream discharge values observed in the study system \citep{Hirsch10}.  Normalized values for pollutant concentration are obtained by averaging the model predictions across the discharge values that are likely to occur on a given day to provide an estimate that is independent of flow variation.  Similarly, predicted values of \ac{DO} concentration were normalized to remove variation from tidal height changes, although the approach differs slightly from \citet{Hirsch10}.  Only two time series are predicted from the observed time series $DO_{obs}$, rather than creating an interpolation grid.  The first time series is predicted values from the model that represent \ac{DO} as a function of time and observed tide.  The second time series are the same predictions conditional on time and constant tidal height set to the mean:
\begin{equation} \label{do_tid}
DO_{tid} = f(DO_{obs}|H, t)
\end{equation}
\begin{equation} \label{do_nrm}
DO_{nrm} = f(DO_{obs}|\bar{H}, t)
\end{equation}
The first time series isolates variation in observed \ac{DO} as a function of physical advection, whereas the second represents variation related to biological processes (\cref{fig:do_dtd}).
